Fluent statistical computing interfaces for biological data analysis

### 2020-10-28 response/response.Rmd

>Pg 4: In Section 1.2, could you add a few more references on existing methods for integration of genomic data and then re-emphasize the point that however existing methods have not thought much about the interoperability between the tidyverse and Bioconductor approaches?

I have expanded this section to reflect these comments.

### Chapter 2 comments

This chapter has already been published and it appears most of the comments pertain to the use and learning of the **plyranges** package. I believe they are best addressed through the **plyranges** documentation and vignette. I have amended the thesis to include the package vignette as an appendix.

@@

### Chapter 3 comments

> Almost all of Chapter 2 discusses the use of GRanges objects. As most of Chapter 3 discusses the use of the SummarizedExperiment object, could you add a motivating figure at the start explaining the relationship between a SummarizedExperiment and GRanges object (i.e. rowRanges)? I do not think this is obvious to a reader who may not very familiar with these objects. However, I think it is an important aspect to explain in this chapter.

I agree and have amended this by adding a figure which describes these features.

> Pg 31: Figure 3.1 is fairly low resolution. Is it possible to make a higher resolution version?

> Pg 31: Figure 3.1 demonstrates how you use plyranges to integrate the results from DE and DA (i.e. finding overlaps). Would it also make sense to the integration step prior to modeling? I am thinking
about this as a way to reduce the amount of tests performed for DE and DA.

I think it would make sense but it is out of scope for this paper.

> Pg 32: The fluentGenomics package needs a reference

>Pg 39: Could you add 1-2 sentences (and maybe a reference) about what is a MA plot and why it’s
useful for visualizing DE results?

> Pg 40: I think it would be useful if you could comment on the results() function a bit more to explain
what is happening under the hood there. It’s a pretty critical step for everything downstream after identifying the DE genes in preparation for the integration step. As it stands, I felt like it was kind of glossed over.

>Pg 45: More motivation could be provided on why you want to generate random samples of size equal to the number of DE-genes from the other_genes with your larger question in mind of “how many DA peaks are near DE genes relative to”other" non-DE genes¿‘. It is stated on pg 46”we minimize the variance on the enrichment statistics induced by the sampling process", but it might be good to explain this in greater detail at a higher level.

I have expanded on this now.

### 2020-10-28 Rmd/03-fluentGenomics.Rmd
finishing touches

```{r workflow, fig.cap = "(ref:workflow)"}
knitr::include_graphics(here::here("img", "workflow-fluentgenomics.pdf"))
```

@@

Throughout the workflow (Figure \@ref(fig:workflow)), we will use existing
Bioconductor infrastructure to understand these datasets. In particular, we
will emphasize the use of the Bioconductor packages **plyranges** and **tximeta**.
The **plyranges** package fluently transforms data tied to genomic ranges using
operations like shifting, window construction, overlap detection, etc. It is
described by @Lee2019 and leverages underlying core Bioconductor infrastructure
[@granges; @bioc] and the **tidyverse** design principles @tidyverse.

@@

First, we specify a directory `dir`, where the quantification files are stored.
You could simply specify this directory with:

```{r dir, eval=FALSE, echo = TRUE}
dir <- "/path/to/quant/files"
```

@@

```{r coldata-rna, echo = TRUE}
library(dplyr)
library(readr)
colfile <- file.path(dir, "coldata.csv")
coldata <- read_csv(colfile) %>%
  select(
    names,
    id = sample_id,
    line = line_id,
    condition = condition_name
  ) %>%
  mutate(
    files = file.path(dir, "quants", names, "quant.sf.gz"),
    line = factor(line),
    condition = relevel(factor(condition), "naive")
  )
coldata
```

@@

```{r linkedtxome-ex, eval = FALSE, echo = TRUE}
makeLinkedTxome(
  indexDir=file.path(dir, "gencode.v29_salmon_0.12.0"),
  source="Gencode",
  organism="Homo sapiens",
  release="29",
  genome="GRCh38",
  fasta="gencode.v29.transcripts.fa.gz", # ftp link to fasta file
  gtf=file.path(dir, "gencode.v29.annotation.gtf.gz"), # local version
  write=FALSE
)
```

@@

To see the results of the expression analysis, we can generate a summary table and a mean-abundance (MA) plot [@Dudoit2002-vz]:

@@

(ref:maplot) Visualization of `DESeq2` results as an "MA plot". Genes  that have an adjusted *p-value* below 0.01 are colored red.In this case the LFC between conditions is shown on the y-axis, while the average normalised gene counts accross all conditions are shown on the x-axis. The assumption of an RNA-seq analysis that most genes are not DE, so most genes are scattered about zero on the y-axis while genes that have evidence of DE are far from the zero baseline. The x-axis gives a sense of the total expression of the gene.

@@

We now wish to extract genes for which there is evidence that the LFC is *not*
large. We perform this test by specifying an LFC threshold and an alternative
hypothesis (`altHypothesis`) that the LFC is less than the threshold in
absolute value. To visualize the result of this test, you can run `results`
without `format="GRanges"`, and pass this object to `plotMA()` as before. We
label these genes as `other_genes` and later as "non-DE genes", for comparison
with our `de_genes` set.

@@

We can repeat this many times to create many samples via `replicate()`. By
replicating the sub-sampling multiple times, we minimize the variance on the
enrichment statistics induced by the sampling process.

>>>>

```{r workflow, fig.cap = "(ref:workflow)"}
knitr::include_graphics(here::here("img", "workflow-fluentgenomics.png"))
```

@@

Throughout the workflow (Figure \@ref(fig:workflow)), we will use existing
Bioconductor infrastructure to understand these datasets. In particular, we
will emphasize the use of the Bioconductor packages **plyranges** and **tximeta** and the _SummarizedExperiment_ class (Figure \@ref(fig:se).
The **plyranges** package fluently transforms data tied to genomic ranges using
operations like shifting, window construction, overlap detection, etc. It is
described by @Lee2019 and leverages underlying core Bioconductor infrastructure
[@granges; @bioc] and the **tidyverse** design principles @tidyverse.

```{r se, fig.cap = "(ref:se)"}
knitr::include_graphics(here::here("img", "se-digram.png"))
```

(ref:se): A _SummarizedExperiment_ orients bulk assay measurements as matrices where rows correspond to features, and columns correspond to samples. The rows have their own accessor functions which contains additional data about the feature of interest, in this `rowRanges()` returns a _GRanges_  where rows are the genomic coordinates for each gene measured in the assay, while `colData()` returns a _DataFrame_  where rows contain information about the samples.

@@

First, we specify a directory `path`, where the quantification files are stored.
You could simply specify this directory with:

```{r dir, eval=FALSE, echo = TRUE}
path <- "/path/to/quant/files"
```

@@

```{r coldata-rna, echo = TRUE}
library(dplyr)
library(readr)
colfile <- file.path(path, "coldata.csv")
coldata <- read_csv(colfile) %>%
  select(
    names,
    id = sample_id,
    line = line_id,
    condition = condition_name
  ) %>%
  mutate(
    files = file.path(path, "quants", names, "quant.sf.gz"),
    line = factor(line),
    condition = relevel(factor(condition), "naive")
  )
coldata
```

@@

```{r linkedtxome-ex, eval = FALSE, echo = TRUE}
makeLinkedTxome(
  indexDir=file.path(path, "gencode.v29_salmon_0.12.0"),
  source="Gencode",
  organism="Homo sapiens",
  release="29",
  genome="GRCh38",
  fasta="gencode.v29.transcripts.fa.gz", # ftp link to fasta file
  gtf=file.path(path, "gencode.v29.annotation.gtf.gz"), # local version
  write=FALSE
)
```

@@

The `results()` function extracts a summary of the DE analysis: in this case for each gene we have the LFC comparing the two cell lines, the Wald test statistic of LFC from the fitted negative binomial GLM, and associated p-value and corrected p-value accounting for the FDR. To see the results of the expression analysis, we can generate a summary table and a mean-abundance (MA) plot [@Dudoit2002-vz]:

@@

(ref:maplot) Visualization of `DESeq2` results as an "MA plot". Genes  that have an adjusted *p-value* below 0.01 are colored red.In this case the LFC between conditions is shown on the y-axis, while the average normalised gene counts across all conditions are shown on the x-axis. The assumption of an RNA-seq analysis that most genes are not DE, so most genes are scattered about zero on the y-axis while genes that have evidence of DE are far from the zero baseline. The x-axis gives a sense of the total expression of the gene.

@@

We now wish to extract genes for which there is evidence that the LFC is *not*
large. We perform this test by specifying an LFC threshold and an alternative
hypothesis (`altHypothesis`) that the LFC is less than the threshold in
absolute value. In this case, the p-values are taken as maximum of the upper and lower Wald tests under the hypothesis absolute value of the estimated LFC is lower than the threshold. To visualize the result of this test, you can run `results`
without `format="GRanges"`, and pass this object to `plotMA()` as before. We
label these genes as `other_genes` and later as "non-DE genes", for comparison
with our `de_genes` set.

@@

We can repeat this many times to create many samples via `replicate()`. We are sub-sampling gene sets without evidence of DE to have the same size as DE gene set, because for each set we want to how different the ATAC peaks around these different sets assuming the number of DE genes is fixed. The sampling of the non-DE genes is done without replacement so each replication produces a different set to compare. By replicating the sub-sampling multiple times, we minimize the variance on the
enrichment statistics induced by the sampling process.

### 2020-10-28 Rmd/07-appendix-A.Rmd
finishing touches

We have two `Ranges` objects. The first contains single nucleotide positions
corresponding to an intensity measurement such as a ChiP-seq experiment, while
the other contains coordinates for two genes of interest.

>>>>

We have two `Ranges` objects. The first contains single nucleotide positions
corresponding to an intensity measurement such as a ChIP-seq experiment, while
the other contains coordinates for two genes of interest.
